Lazy Linearity for a Core Functional Language

### 2025-10-17 popl26.tex
End S2

In Core, cases are of the form $\ccase{e_s}{z~\{\ov{\rho_i \to e_i}\}}$,
where $e_s$ is the case \emph{scrutinee}, $z$ is the case \emph{binder}, and
$\ov{\rho_i \to e_i}$ are the case \emph{alternatives}, composed of a pattern
$\rho_i$ and of the corresponding expression $e_i$. 
%\begin{enumerate}
%\item
The case scrutinee is always evaluated to Weak Head Normal Form (WHNF) and
%
% \item
% evaluating an expression that is already in WHNF is a no-op, that is,
% no computation occurs; 
%
the case binder is an alias to the result of evaluating the scrutinee
to WHNF. Moreover, in Core, the alternative patterns are always exhaustive.
% i.e. there always exists
% a pattern that matches the WHNF of a value resulting from evaluating the
% scrutinee, where a pattern is either a wildcard that matches all expressions
% ($\_$), or a constructor and its linear and non-linear component binders
% ($K~\ov{x}\ov{y}$, with $\ov{x}$ as linearly-bound variables and $\ov{y}$ as
% unrestricted ones).
%
%\end{enumerate}
%
% ROMES:TODO: We should talk about WHNF in the background, not here.
% \parawith{WHNF} An expression is in Weak Head Normal Form when 
%
\ROUNDTWO{The first} example program constructs a tuple from linear resources,
matches on it, then uses both linearly-bound variables from the tuple pattern.
This is well-typed in Linear Haskell:
\begin{working}
\begin{lstlisting}
f6 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* c) -> c
f6 x y use = case (x,y) of z { (a,b) -> use a b }
\end{lstlisting}
\end{working}
\ROUNDTWO{Perhaps surprisingly}, a similar program which discards the pattern
variables, and instead uses the resources \ROUNDTWO{from} the scrutinee
\ROUNDTWO{again}, is still linear, despite not being accepted by Linear Haskell:
\begin{notyet}
\begin{lstlisting}
f7 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* c) -> c
f7 x y use = case (x,y) of z { (a,b) -> use x y }
\end{lstlisting}
\end{notyet}
\incode{f7} \ROUNDTWO{can be seen as} linear by \ROUNDTWO{realizing} that
the tuple being scrutinized is already in WHNF, so evaluating it again does
not consume either $x$ \ROUNDTWO{nor} $y$. Even if the tuple \ROUNDTWO{paired}
two \ROUNDTWO{arbitrary} expressions which \ROUNDTWO{captured} $x$ and $y$,
\ROUNDTWO{because $a$ and $b$ are not forced in the continuation, $x$ and $y$ would remain unused.}
\ROUNDTWO{On the other hand, as seen in the next example}, if $a$ were used in 
the case body, then $x$ \ROUNDTWO{is consumed via $a$ and must not be used again.}
Thus, \incode{f8} must be rejected:
\begin{noway}
\begin{lstlisting}
f8 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* a *' $\lolli$ '* c) -> c
f8 x y use = case (x,y) of z { (a,b) -> use a x }
\end{lstlisting}
\end{noway}
This idea that $x$ and $a$ are mutually exclusive is \ROUNDTWO{akin to how} let
bindings \ROUNDTWO{are} mutually exclusive \ROUNDTWO{with the set} of resources
\ROUNDTWO{they capture, in the let-continuation}.
\ROUNDTWO{Forcing a pattern variable evaluates the expression for the
corresponding constructor field, and all linear variables captured by that
expression may end up being used}.
% If we do not use the pattern variables, then we do not run the computation and
% thus no resources are consumed.
%
A third (semantically well-typed, but rejected by Linear Haskell) option is to use the case binder $z$ instead of $a,b$ or $x,y$:
\begin{notyet}
\begin{lstlisting}
f9 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* c) -> c
f9 x y use = case (x,y) of z { (a,b) -> uncurry use z }
\end{lstlisting}
\end{notyet}
Again, $z$ is mutually exclusive with $(a,b)$ and with $(x,y)$, but at least
one of the three must \ROUNDTWO{be used} to ensure the
linear resources are consumed. Essentially, in this example, using $a$ entails
using the resource $x$, $b$ the resource $y$, and the case binder $z$ entails
using both $a$ and $b$.

Dually, consider the scrutinee to be an expression that's not in WHNF, s.t.
evaluating it to WHNF will require performing computation and thus consume linear
resources (\incode{f10} below is well-typed in our system but rejected
by Linear Haskell):
\begin{notyet}
\begin{lstlisting}
f10 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* (c,d)) -> (c,d)
f10 x y use = case use x y of z { (a,b) -> z }
\end{lstlisting}
\end{notyet}
Unlike when the scrutinee was in WHNF, we can no longer use $x,y$ in the case
alternatives, but we must still use either the case binder $z$ or the linear
pattern variables $a,b$. %  e.g. it would be quite disastrous if any of the following typechecked:
% \begin{noway}
% \begin{lstlisting}
% doubleFree :: Ptr *' $\lolli$ '* (Ptr *' $\lolli$ '* Result) -> Result
% doubleFree x free = case free x of z { Result v -> free x }
% \end{lstlisting}
% \end{noway}
% % Remover
% \begin{noway}
% \begin{lstlisting}
% leakPointer :: Ptr -> ()
% leakPointer x = case id x of z { _ -> () }
% \end{lstlisting}
% \end{noway}
%
%whereas this is fine:
%\begin{notyet}
%\begin{code}
%f16 :: Ptr -o (Ptr -o (Value,Int)) -> (Value,Int)
%f16 x free = case K (free x) of z { K y -> free x }
%\end{code}
%\end{notyet}
%
The result of evaluating the scrutinee must be consumed exactly once to guarantee
that the resources used in the scrutinee are fully consumed. %, or
                                %risk them being only ``almost''
                                %consumed.
For instance, take
\incode{use} in \incode{f10} to be the pairing function \incode{(,)}: it is not sufficient for \incode{use x y} to be
evaluated to WHNF to consume \incode{x} and \incode{y}. Otherwise, if all the resources were
considered to be fully consumed after the scrutinee were evaluated in a case
expression, we could simply ignore the pattern variables, potentially
discarding linear resources. In
short, if the scrutinee is not in WHNF we must either consume the case binder
or the linear components of the pattern.

On the other hand, if we match on a constructor without any linear components,
all linear resources used in the scrutinee must have been fully consumed in
that branch since the result of evaluating it can be consumed unrestrictedly
(by definition, a constructor application to unrestricted arguments is also
unrestricted).
%
Hence, in the branch for a constructor without linear fields, the case
binder can also be used unrestrictedly. For example, the program in
Figure~\ref{fig:nolin} is semantically linear.
%
\begin{figure}[h]
 \begin{minipage}{0.5\textwidth}
\begin{notyet}
\begin{lstlisting}
f11 :: () *' $\lolli$ '* ()
f11 x = case x of z { () -> z <> z }
\end{lstlisting}
\end{notyet}
\vspace{-0.5cm}
\caption{No Linear Fields\label{fig:nolin}}
\end{minipage}%
\begin{minipage}{0.5\textwidth}
\begin{limitation}
\begin{lstlisting}
f12 :: a *' $\lolli$ '* a
f12 x = case K1 x of z { K2 -> x; K1 a -> x }
\end{lstlisting}
\end{limitation}
\vspace{-0.5cm}
\caption{Absurd Branches\label{fig:absurd}}
\end{minipage}
\end{figure}
%
A second example of an unrestricted pattern, where \incode{K2} has no fields
and \incode{K1} has one linear field, is shown in Figure~\ref{fig:absurd}.
The use of \incode{x} in the \incode{K2} branch is valid despite \incode{x}
also occurring in the scrutinee because this branch is never taken. In fact,
any arbitrary resource could be used in absurd branches and be understood as
linear (despite not being seen as such by our system).

>>>>

In Core, cases are of the form $\ccase{e_s}{z~\{\ov{\rho_i \to e_i}\}}$,
where $e_s$ is the case \emph{scrutinee}, $z$ is the case \emph{binder}, and
$\ov{\rho_i \to e_i}$ are the case \emph{alternatives}, composed of a pattern
$\rho_i$ and of the corresponding expression $e_i$. 
%\begin{enumerate}
%\item
The case scrutinee is always evaluated to Weak Head Normal Form (WHNF) and
%
% \item
% evaluating an expression that is already in WHNF is a no-op, that is,
% no computation occurs; 
%
the case binder is an alias to the result of evaluating the scrutinee
to WHNF. Moreover, in Core, the alternative patterns are always exhaustive.
% i.e. there always exists
% a pattern that matches the WHNF of a value resulting from evaluating the
% scrutinee, where a pattern is either a wildcard that matches all expressions
% ($\_$), or a constructor and its linear and non-linear component binders
% ($K~\ov{x}\ov{y}$, with $\ov{x}$ as linearly-bound variables and $\ov{y}$ as
% unrestricted ones).
%
%\end{enumerate}
%
% ROMES:TODO: We should talk about WHNF in the background, not here.
% \parawith{WHNF} An expression is in Weak Head Normal Form when 
%
\ROUNDTWO{The first} example program constructs a tuple from linear resources,
matches on it, then uses both linearly-bound variables from the tuple pattern.
This is well-typed in Linear Haskell:
\begin{working}
\begin{lstlisting}
f6 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* c) -> c
f6 x y use = case (x,y) of z { (a,b) -> use a b }
\end{lstlisting}
\end{working}
\ROUNDTWO{Perhaps surprisingly}, a similar program which discards the pattern
variables, and instead uses the resources \ROUNDTWO{from} the scrutinee
\ROUNDTWO{again}, is still linear, despite not being accepted by Linear Haskell:
\begin{notyet}
\begin{lstlisting}
f7 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* c) -> c
f7 x y use = case (x,y) of z { (a,b) -> use x y }
\end{lstlisting}
\end{notyet}
\incode{f7} \ROUNDTWO{can be seen as} linear by \ROUNDTWO{realizing} that
the tuple being scrutinized is already in WHNF, so evaluating it again does
not consume either $x$ \ROUNDTWO{nor} $y$. Even if the tuple \ROUNDTWO{paired}
two \ROUNDTWO{arbitrary} expressions which \ROUNDTWO{captured} $x$ and $y$,
\ROUNDTWO{because $a$ and $b$ are not forced in the continuation, $x$ and $y$ would remain unused.}
\ROUNDTWO{On the other hand, as seen in the next example}, if $a$ were used in 
the case body, then $x$ \ROUNDTWO{is consumed via $a$ and must not be used again.}
Thus, \incode{f8} must be rejected:
\begin{noway}
\begin{lstlisting}
f8 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* a *' $\lolli$ '* c) -> c
f8 x y use = case (x,y) of z { (a,b) -> use a x }
\end{lstlisting}
\end{noway}
This idea that $x$ and $a$ are mutually exclusive is \ROUNDTWO{akin to how} let
bindings \ROUNDTWO{are} mutually exclusive \ROUNDTWO{with the set} of resources
\ROUNDTWO{they capture, in the let-continuation}.
\ROUNDTWO{Forcing a pattern variable evaluates the expression for the
corresponding constructor field, and all linear variables captured by that
expression may end up being used}.
% If we do not use the pattern variables, then we do not run the computation and
% thus no resources are consumed.
%
A third option, semantically \ROUNDTWO{linear}, but rejected by Linear Haskell, is to use
the case binder $z$ instead of $a,b$ or $x,y$:
\begin{notyet}
\begin{lstlisting}
f9 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* c) -> c
f9 x y use = case (x,y) of z { (a,b) -> uncurry use z }
\end{lstlisting}
\end{notyet}
Again, $z$ is mutually exclusive with $(a,b)$ and with $(x,y)$, but at least
one of the three must \ROUNDTWO{be used} to ensure the
linear resources are consumed. Essentially, in this example, using $a$ entails
using the resource $x$, $b$ the resource $y$, and the case binder $z$ entails
using both $a$ and $b$.

\ROUNDTWO{Conversely}, if the scrutinee is an expression that's not in WHNF,
evaluation to WHNF \ROUNDTWO{may indeed} consume \ROUNDTWO{some or all of the}
linear resources \ROUNDTWO{captured by said scrutinee}. \ROUNDTWO{Consider the following
non-WHNF scrutinee example},
\incode{f10}, \ROUNDTWO{which is accepted} by our system:
\begin{notyet}
\begin{lstlisting}
f10 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* (c,d)) -> (c,d)
f10 x y use = case use x y of z { (a,b) -> z }
\end{lstlisting}
\end{notyet}
\ROUNDTWO{In contrast to} a scrutinee in WHNF, \ROUNDTWO{the resources captured
by a non-WHNF scrutinee} (here, $x,y$), \ROUNDTWO{conservatively, can no longer
be used} in the case alternatives, \ROUNDTWO{since reducing \incode{use x y} to
WHNF may consume (parts of) $x$ and/or $y$}. \ROUNDTWO{Consequently}, either the case
binder $z$ or the linear pattern variables $a,b$ must be used exactly
once,
%  e.g. it would be quite disastrous if any of the following typechecked:
% \begin{noway}
% \begin{lstlisting}
% doubleFree :: Ptr *' $\lolli$ '* (Ptr *' $\lolli$ '* Result) -> Result
% doubleFree x free = case free x of z { Result v -> free x }
% \end{lstlisting}
% \end{noway}
% % Remover
% \begin{noway}
% \begin{lstlisting}
% leakPointer :: Ptr -> ()
% leakPointer x = case id x of z { _ -> () }
% \end{lstlisting}
% \end{noway}
%
%whereas this is fine:
%\begin{notyet}
%\begin{code}
%f16 :: Ptr -o (Ptr -o (Value,Int)) -> (Value,Int)
%f16 x free = case K (free x) of z { K y -> free x }
%\end{code}
%\end{notyet}
%
since the linear resources in the scrutinee are \ROUNDTWO{only} fully consumed
if the result of evaluating the scrutinee \ROUNDTWO{is also} fully consumed . %, or
                                %risk them being only ``almost''
                                %consumed.
For instance, take
\incode{use} in \incode{f10} to be the pairing function \incode{(,)}: it is not
sufficient for \incode{use x y} to be evaluated to WHNF to consume \incode{x}
and \incode{y}.
%
If all the resources were considered to be fully consumed after
\ROUNDTWO{evaluating the} scrutinee to \ROUNDTWO{WHNF}, the pattern variables
could simply be ignored and thus linear resources potentially discarded.
In \ROUNDTWO{practice}, if the scrutinee is not in WHNF then \ROUNDTWO{either}
the case binder or the linear components of the pattern \ROUNDTWO{must be consumed}.

\ROUNDTWO{Alternatively}, if a constructor without any linear components \ROUNDTWO{is matched},
all linear resources used in the scrutinee must \ROUNDTWO{already} have been
fully consumed in that branch since the result of evaluating it can be consumed
unrestrictedly (by definition, a constructor application to unrestricted
arguments is also unrestricted).
%
Hence, in the \ROUNDTWO{case-}branch for a constructor without linear fields, the case
binder can also be used unrestrictedly. For example, the program in
Figure~\ref{fig:nolin} is semantically linear.
%
\begin{figure}[h]
 \begin{minipage}{0.5\textwidth}
\begin{notyet}
\begin{lstlisting}
f11 :: () *' $\lolli$ '* ()
f11 x = case x of z { () -> z <> z }
\end{lstlisting}
\end{notyet}
\vspace{-0.5cm}
\caption{No Linear Fields\label{fig:nolin}}
\end{minipage}%
\begin{minipage}{0.5\textwidth}
\begin{limitation}
\begin{lstlisting}
f12 :: a *' $\lolli$ '* a
f12 x = case K1 x of z { K2 -> x; K1 a -> x }
\end{lstlisting}
\end{limitation}
\vspace{-0.5cm}
\caption{Absurd Branches\label{fig:absurd}}
\end{minipage}
\end{figure}
%
A second example of an unrestricted pattern, where \incode{K2} has no fields
and \incode{K1} has one linear field, is shown in Figure~\ref{fig:absurd}.
The use of \incode{x} in the \incode{K2} branch is valid despite \incode{x}
also occurring in the scrutinee because this branch is never taken. In fact,
any arbitrary resource could be used in absurd branches and be understood as
linear (despite not being seen as such by our system).
